---
title: "aRt of the figure:<br>explore and visualize your data using R"
output: 
  html_document: 
    toc: yes
    theme: spacelab
---
_[Andres Karjus](http://andreskarjus.github.io)_

<div style="text-align: justify"> 

This functions as a cover page for all the *aRt of the Figure* workshops. Each iteration has had a slightly different structure, length and topic coverage. If you have signed up to one of the workshops, you should have received an email ahead of time with instructions. If you are here because you signed up but did not get an email, contact the organizers to make sure you are registered.

**Upcoming** (watch this space)

- Introduction to data exploration and visualization with R. Workshop at the PPLS LEL Spring Retreat 2019 to Firbush | [Installation instructions](https://andreskarjus.github.io/artofthefigure/preworkshop_firbush2.html)  | <a download href="https://raw.githubusercontent.com/andreskarjus/artofthefigure/master/artofthefigure_firbush2.Rmd">script</a>
- Attractive, Interactive, Ready for the Web: Visualizing Your Data Using R. Workshop at the International Corpus Linguistics Conference 2019, Cardiff, July 2019 | [overview](http://www.cl2019.org/wp-content/uploads/2019/02/visualizing_data_with_R_cl2019.pdf) | (more info soon!)
- Attractive, Interactive, Ready for the Web: Visualizing Your Data Using R. Workshop at the Digital Humanities 2019 conference, Utrecht, July 2019.
- Short workshop at the Scots Syntax Atlas Data Hack, content tbd. June 2019.
- Jupyter Notebooks taster, SGSSS Summer School 2019 (co-teaching; not R; but will include examples from the inter-language plotly package which has been used here a lot, see examples below)

**Past**

- Data visualization with R for corpus linguists, a workshop for Corpus Linguistics in Scotland, 12 April 2019 | <a download href="https://raw.githubusercontent.com/andreskarjus/artofthefigure/master/artofthefigure_clis.Rmd">download the script</a> | [Installation instructions and preparatory tutorial](https://andreskarjus.github.io/artofthefigure/preworkshop_clis.html) 
- Centre for Language Evolution PhD Retreat, January 2019 |  <a download href="https://raw.githubusercontent.com/andreskarjus/artofthefigure/master/artofthefigure_clephd.Rmd">the script</a>
- PPLS Writing Centre summer quantitative workshops series 2018
- SGSSS Summer School 2018
- PPLS LEL Spring Retreat 2018
- PPLS Writing Centre short workshops for 2017-2018

Note that if the script opens as a plain text file in your browser, you'll need to right-click it and choose "Save as..."; and make sure the saved file has the `.Rmd` extension, not a `.txt` extension (fix it if needed). 
<br>The short intro motivational "why-use-R" slides are [here](https://andreskarjus.github.io/artofthefigure/intro). 
<br>It's all open source: the entire Github repo with additional info can be found [here](https://github.com/andreskarjus/artofthefigure). 



---

Below is a little showcase of the kinds of plots and figures you could learn to create if you participated in one of the workshops, from very simple base graphics plots to complex networks, interactive plots and zoomable maps (content depending on the length and focus of a given workshop. <br> Check [my website](https://andreskarjus.github.io/) for more details on consulting, bookings and past workshops. If you like the materials and would like to invite me to come and teach you and/or your students how to do all this (in an easy-to-learn, efficient and reproducible manner), feel free to get in touch to discuss options.

</div> 

```{r,echo=F}
htmltools::includeHTML("plots_fragment.html") 
```


```{r, echo=F, message=F, warning=F, out.width = '100%'}
library("quanteda", quietly=T, warn.conflicts=F) 
library("plotly", quietly=T, warn.conflicts=F)
library(RColorBrewer)
library(scales)
library(magrittr)
library(igraph, quietly = T)

# A look into the usage of some words across centuries
termmat_prop = dfm_weight(dfm(data_corpus_inaugural, tolower = T, stem=F,remove=stopwords(), remove_punct=T), "prop") # use normalized frequencies
words = c("america", "states", "americans", "white", "black",  "dream", "hope", "happiness",  "business", "dollars","money",  "immigrants", "peace", "war", "terror")

words2 = c("america$", "states", "american", "nation", "white", "black",  "dream", "hope", "great", "happiness",  "business", "dollar","money",  "immigra", "peace", "war", "army", "terror", "god", "lord")
newmat = round(sapply(words2, function(x) rowSums(termmat_prop[, grep(x, colnames(termmat_prop))])),5)

p = plot_ly(x=words2, y=rownames(newmat), z=newmat, type="heatmap", colors = colorRamp(c("white", "lightgray", "gray", "darkgray", "black")),showscale = F)
p = layout(p,yaxis=list(tickfont=list(size=6)), xaxis=list(tickfont=list(size=6)),  margin = list(l=130, b=70), paper_bgcolor=rgb(0.95,0.95,0.95)) %>% config(displayModeBar = F)
p


```


```{r, echo=F,message=F, warning=F,out.width = '100%', out.height=600}
# <div style="float: left; width: 100%;height: 400px;">
# </div>
load("/Users/pplsuser/Dropbox/6petamine/artofthefigure_all/gl.RData")
gl %>% config(displayModeBar = F)
```


<div style="float: left; width: 50%; height: 400px;">
```{r, echo=F,message=F, warning=F, fig.height=4,  out.width = '100%'}
# library("raster", quietly = T)
# uk = getData("GADM", country = "United Kingdom", level = 2) # download UK map (needs the raster package to be loaded)
# plot(uk, lwd=0.4, border=rgb(0,0,0,0.2)) # plot the UK
# grid()            # add grid

library(plotly, quietly = T)
df=data.frame(d=c(runif(100,-1,1)+rnorm(100,0,0.5),sort(runif(100,-0.6,0.6))+rnorm(100,0,0.5), sort(runif(100,-1,1))+rnorm(100, 0, 0.1)), R2=as.factor(c(rep("0",100), rep("0.3",100), rep("0.95",100) )))
retimes = data.frame(cbind(subj1=c(runif(20, 0,0.1), seq(0.1, 2.8,length.out = 80)) *  runif(100, 0.7, 1.1 ), subj2=sort(rnorm(100, 1,0.3))*runif(100,0.4,0.8), subj3=sort(rnorm(100, 1,0.3))*runif(100,0.9,1.1) ))

p4= plot_ly(type="scatter", data = retimes, mode="none")  %>% 
  add_trace(y=~subj1, mode="lines", name="Subject 1", color=I("darkgray")) %>% 
  add_trace(y=~subj2, mode="lines", name="Subject 2", color=I("darkred")) %>% 
  layout(yaxis=list(title="Reaction time"), xaxis=list(title="Experiment progress"), showlegend=F) %>% config(displayModeBar = F)

cols=viridis_pal(alpha=1,begin=0.3, end=0.8)(3)
p2 = plot_ly(x= iris$Petal.Length,y=iris$Sepal.Width, z=iris$Sepal.Length, type="scatter3d",mode="markers",color=iris$Species, colors=cols, opacity=0.4, marker=list(size=7)) %>% layout(showlegend=F) %>% config(displayModeBar = F)


  n=100 # number of iterations
  m1=data.frame(x=rep(1,n), y=rep(1,n))
  m2=data.frame(x=rep(1,n), y=rep(1,n))
  for(i in 2:n){
    s=seq(m1$x[i-1]-3,m1$x[i-1]+3, 0.5); s=s[s>0]
    m1$x[i]=sample(s,1, prob=abs(s))
    s=seq(m1$y[i-1]-3,m1$y[i-1]+3, 0.5); s=s[s>0]
    m1$y[i]=sample(s,1,  prob=abs(s))
    s=seq(m2$x[i-1]-2,m2$x[i-1]+2, 0.5); s=s[s>0]
    m2$x[i]=sample(s,1, prob=rev(abs(s-m1$x[i])) )
    s=seq(m2$y[i-1]-2,m2$y[i-1]+2, 0.5); s=s[s>0]
    m2$y[i]=sample(s,1, prob=rev(abs(s-m1$y[i])) )
  }
  m=cbind(rbind(m1,m2), step=c(1:n,1:n), agent=rep(c("a", "b"), each=n))




suppressWarnings({
p1 = plot_ly(df, y=~d, x=rep(seq(-1,1,length.out = 100),3), color=~R2, colors=c("dimgray", "darkgreen", "darkred"), hoverinfo="none", frame = ~R2,type = 'scatter', mode = 'markers',showlegend = F) %>%
  layout(yaxis=list(title="")) %>% 
  animation_button(visible=FALSE) %>% config(displayModeBar = F)
p1
})
```
</div>
<div style="float: right; width: 50%; height: 400px;">
```{r, echo=F,message=F, warning=F, fig.height=4,  out.width = '100%'}
p12 = plot_ly(m, x=~x, y=~y, type = 'scatter', mode = 'markers', 
        frame=~step, color=~agent, marker=list(size=20), colors=c(rgb(0,0,0.5,0.8), rgb(0.5,0,0,0.8))) %>% 
  layout(showlegend = FALSE,yaxis=list(title=""),xaxis=list(title="")) %>% 
  animation_opts(frame = 200) %>% config(displayModeBar = F)
#subplot(p1, p12, shareX = F, shareY = F,which_layout = "merge") %>% layout(scene = list(domain = list(x = c(0.6, 1.2), y = c(0,1)))) %>% animation_slider()
p12

```
</div>
<div style="display: inline-block; position: relative; width: 100%;height: 400px;">
```{r, echo=F,message=F, warning=F, fig.height=3,  out.width = '100%'}
subplot(p4, p2) %>% layout(scene = list(domain = list(x = c(0.6, 1.2), y = c(0,1))))

```
</div>


```{r, echo=F,message=F, warning=F, eval=F}
<div style="float: left; width: 50%; height: 400px;">
eur = data.frame(country = c("AUT","BEL","BGR","HRV","CYP","CZE","DNK","EST","FIN","FRA","DEU","GRC","HUN","IRL","ITA","LVA","LTU","LUX","MLT","NLD","POL","PRT","ROU","SVK","SVN","ESP","SWE","GBR", "NOR", "ISL"), value = sample(seq(0,5,0.1),30)) # create some data

plot_geo(eur, color = ~value, colors = c("gray90", "gray40")) %>% add_trace(z = ~value, locations = ~country, color = ~value) %>% 
  layout(margin=list(b=1,l=1,t=1,r=1, pad=0), geo = list(scope="europe",lonaxis = list(range = c(-10, 28)),lataxis = list(range = c(30, 73)) ), margin=list(l=0,r=0,b=0,t=0))  %>%  hide_colorbar() %>%
  colorbar(title = "", thickness=5) %>% config(displayModeBar = F)
#</div>
```

<div style="display:flex;height:300px">
<div style="flex:1;padding-right:5px">
```{r, echo=F, message=F, warning=F, out.width = '100%'}

library(visNetwork, quietly=T, verbose = F)
library("igraph", quietly = T)  
scots=c("Iain","Irvine","James","Muir","Mungo","Owen","Raibert", "Lyall", "Margaret", "Mairi", "Morag", "Murdina","Rhona", "Sorcha", "Thomasina","Una")
nscots = length(scots) # record the number of people in an object
mates = matrix(sample(0:1,nscots^2,T,prob=c(0.8,0.2)), ncol=nscots, nrow=nscots, dimnames=list(scots, scots)) # this creates a randomized matrix signifying friendships; no need to think about this too hard for now
g = graph_from_adjacency_matrix(mates, mode = "undirected", diag=FALSE) # creates a graph object; this need to have igraph loaded to work
gender = c(rep("m", nscots-9), rep("f", 9)) # create a vector of gender labels for the names
#plot(g, vertex.size=4, vertex.color="lightgray", vertex.frame.color=NA, vertex.label.cex=0.7, vertex.label.color=ifelse(gender=="m", yes="blue",no="tomato"), vertex.label.dist=0.1, vertex.label.font=2, edge.color=rgb(0,0,0,0.3))
mfcolors = ifelse(gender=="m", yes=rgb(0.5,0.5,0.7),no=rgb(0.7,0.5,0.5))
# Let's try something else. Using the same graph data, we'll recreate it using another package, visNetwork, which makes graphs interactable.

scotgraph_v = toVisNetworkData(g) # converts the previous igraph object into a visNetwork object
scotgraph_v$nodes$color = mfcolors # let's use the same node colors
scotgraph_v$nodes$frame = "white"
scotgraph_v$edges$color = "lightgray"
v = visNetwork(nodes = scotgraph_v$nodes, edges = scotgraph_v$edges, height = "500px")
v = visNodes(v, size = 10,  shadow=T, font=list(size = 10))
v = visPhysics(v, enabled=T, solver = "forceAtlas2Based", forceAtlas2Based = list(gravitationalConstant = -10)) 
v 

```
</div>

<div style="flex:1;padding-left:5px">
```{r, echo=F, message=F, warning=F, out.width = '100%'}
library(visNetwork, quietly=T, verbose = F)

speeches = data_corpus_inaugural$documents$texts # extract speeches data from the internal object
speeches = gsub("Washington DC", "DC", speeches) # replace city name to avoid confusion with president Washington (hopefully)
speechgivers = data_corpus_inaugural$documents$President # names of presidents giving the speech
presidents = unique(data_corpus_inaugural$documents$President) # presidents (some were elected more than once)
mentions = matrix(0, ncol=length(presidents), nrow=length(presidents), dimnames=list(presidents, presidents))
for(president in presidents){
  foundmentions = grep(president, speeches)
  mentions[speechgivers[foundmentions], president ] = 1
}

pgraph = graph_from_adjacency_matrix(mentions, mode="directed") # this uses igraph
pgraph_v = toVisNetworkData(pgraph )
v = visNetwork(nodes = pgraph_v$nodes, edges = pgraph_v$edges)
v = visNodes(v, size = 10,  shadow=T, font=list(size = 25), color="gray")
v = visIgraphLayout(v, "layout_in_circle", smooth=T)
v = visEdges(v, arrows = "to",  shadow=T, smooth=list(type="discrete"), selectionWidth=5)
v = visOptions(v, highlightNearest = list(enabled = T, hover = T, degree=1, labelOnly=F, algorithm="hierarchical"), nodesIdSelection = F)
v
```
</div>

</div>

<br>
<br>

<br>