---
title: "SCOSYA Data Hack extra scripts"
author: "Andres Karjus"
date: "11/06/2019"
output: html_document
---


```{r, warning=F, message=F}
# Required packages
library(dplyr)     # for filter() and pipes
library(reshape2)  # for melt()
library(tidyr)     # for separate(); install.packages("tidyr") if necessary 
library(tibble)    # for rownames_to_column (should have been installed with dplyr)
```

```{r}
# Define file path:
PATH = "pathto/file.csv" # file path to SCOSYAdatahackjudgments.csv file
```

```{r, echo=F}
# Define file path:
PATH = "/Users/pplsuser/Downloads/SCOSYAdatahackjudgments.csv" # file path to SCOSYAdatahackjudgments.csv file
```


# Import data, replacing empty cells with NA, assign unique IDs to people, and convert wide to long format:

```{r}
coln = read.csv(PATH, header = F, nrows=2, stringsAsFactors = F) %>% rbind(-3:ncol(.)) %>% apply(2,paste, collapse="_"); coln[1:4] = c("FEATURE", "CODE", "EXAMPLE", "GROUP")
dat_wide = read.csv(PATH, na.strings=c(""), skip=2, header = F, col.names = coln)
dat_long = melt(dat_wide, measure.vars = 5:ncol(dat_wide), value.name="rating") %>% separate(variable, c("agegroup", "location", "id"),  sep = "_"); rm(coln)
```


```{r}
# have a look:
dat_wide[1:4,1:7]
head(dat_long) 
```



# Optional: add coordinates to each data point in dat_long
To make matching possible, convert spaces and ' into periods (column names can't have spaces, so they get replaces with .)

```{r}
locs = read.csv("http://scotssyntaxatlas.ac.uk/api/v1/csv/locations", skip=1, stringsAsFactors = F) %>% mutate(display_town=gsub("[' ]",".", display_town))
dat_long = cbind(
  dat_long,
  locs[match(dat_long$location, locs$display_town), c("display_lat", "display_lng")]
)

head(dat_long) # have a look
```



# Examples

## Usage of the long format data

```{r}
library(ggplot2)
ggplot(dat_long %>% filter(CODE=="A1"), 
       aes(y=display_lat, x=display_lng, color=agegroup )) + 
  geom_point()
```


## Examples of usage of the wide format data

```{r}
library(plotly)  # for ggplotly()
distmat = dist(t(dat_wide[ ,-c(1:4)]) ) 
# dist() calculates euclidean distances between rows, hence the t() transpose

# simple dimension reduction using multidimensional scaling; separating id tags for plotting purposes
mds = cmdscale(distmat, k=2) %>% as.data.frame() %>% tibble::rownames_to_column("tags") %>% separate("tags", c("agegroup", "location", "id"),  sep = "_") 

{ ggplot(mds, aes(y=V2, x=V1, color=agegroup, group=location)) + 
    geom_point(size=1, alpha=0.6) + 
    labs(x="MDS dimension 1", y="MDS dimension 2") +
    theme_bw()
  } %>% ggplotly()
```

Click on legend items to hide/show traces!


## Adding longitude info as color
This just cuts the available longitudes into three equally spaced groups, [-7.29,-5.18] (-5.18,-3.08] (-3.08,-0.975] (but there will be different number of people in each group)

```{r}
mds$display_lng = 
  cut(locs[match(mds$location, locs$display_town),  "display_lng"],
      breaks = 3, include.lowest = T, labels = c("west", "mid", "east")
      )

{ ggplot(mds, aes(y=V2, x=V1, shape=agegroup, color=display_lng, group=location)) + 
    geom_point(size=2, alpha=0.7, position="jitter") + 
    labs(x="MDS dimension 1", y="MDS dimension 2", title="Coloured by longitude (west-east)") +
    scale_shape_manual(values=c(1,4)) + 
    scale_color_manual(values=c("navy", "darkgray", "tomato")) +
    theme_bw()
} %>% ggplotly()
```

Click on legend items to hide/show traces!


